Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework

### 2016-07-19 ms/ms.Rnw
Added references

\dropcap{R}ates of plant mortality are known to be highly variable among individuals, among coexisting species, between forests and from year-to-year [refs]. This variation has considerable consequences for forest structure and dynamics. Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools (Chave et al 200X), while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun. In models of forest dynamics, variation in mortality rates has been found to have a larger impact on forest structure than absolute growth rates \cite{Johnson-2016}. As such, improving our ability to understand the mortality process is therefore a priority for making skilful predictions about forest dynamics and carbon, water or nutrient cycles; especially in an era of rapid environmental change.

Two substantial challenges exist when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates (Poorter et al 2008). This requirement arises in part because the observable outcome -- alive vs. dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate). In classic survival analysis \cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$ \cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
\end{equation}
Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required. As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys in long-term monitoring plots containining thousands of individuals and spanning several decades \cite{Anderson-Teixeira-2015a}.

The other major challenge when studying mortality -- similar to that faced in other fields like medicine and engineering -- is to derive suitable shapes for the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality. Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate, traits, and size (XX). Generally, however, these influences have not been formalised as hazard functions (XX). Yet such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function (XX). This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.

A particular challenge is to integrate the observed conflicting relationships between mortality and growth observed within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality than individuals from slower growing species \cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

@@

The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards (Moorcroft, Sortie,McDowell 2011). We show that regardless of growth measure, incorporating both hazards significantly improves model predictive accuracy. This is because the additive effects of both hazards explicitly allow models to account for two independent causes of plant mortality. The growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stressors (e.g. competition, parasitization, localized herbivory). By contrast, the growth-independent hazard accounts for deaths caused by stochastic events that arise irrespective of an individuals growth rate (e.g. windthrow).

Another advantage of partitioning mortality into growth-depedent and growth-independent hazards was that we could estimate the proportion of variation each attributed. Like many other studies (Denslow JS (1987) King DA, Davies SJ, Supardi MNN, Tan S (2005), Ruger 2011), our analyses highlight the importance of light competition as a dominant factor influence tropical tree demographic rates. Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted varability in mortality rates (Fig 5), and thus, suggests that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.

Incorporating the effect of wood density on mortality rates also improved predictive performance. Our analyses revealed that the most parsimonious combination of wood density effects was when it was attributed to only growth-independent mortality. Specifically, we found that high wood density species had lower growth-independent mortality rates relative to their low wood density counterparts. This finding corroborating the observed negative correlation between mortality and wood density reported elsewhere \cite{Kraft:2010kq,Chave:2009iy}. More importantly, however, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.

While wood density effects are now being incorporated in mortality algorithms of many vegetation models (REFS), our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation. This means that such models are likely to severely underestimate the true varability in mortality rates. Consequently, because variation in mortality rates are known to have significant impacts on forest dynamics \cite{Johnson-2016}, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics, with flow on effects for ecosystem proporties such as GPP and NPP.

@@

The underestimation of inter-specific variation suggests that there are other species attributes besides wood density that affect tree mortality rates. Poorter et al. (2006) suggested that a species light demand and maximum size (a correlate to species longevity; Lieberman et al 1985) are two major axes of inter-specific variation in tropical rainforests. While we did not explicitly model mortality as a function of these traits, our posthoc correlations (FIG) revealed that some inter-specific variation was indeed correlated with light demand, but not maximum size. We found that light demanding species (i.e. those with high gap index) were more susceptible to death due to both low growth and stochastic chance and required faster growth rates to achieve a given mortality rate relative to a species that readily recruits in shaded conditions.

We recommend research should focus on examining how other traits influence growth-dependent and growth-independent hazards and what combination of traits minimises unexplained species variability. Researchers, however, should be mindful that the inclusion of additional traits comes at the cost of added complexity, increased uncertainty (i.e. more parameters to estimate) and increased minimum knowledge; all of which diminish the interpretability of models, but also the applicability of predicting mortality rates in many data poor ecosystems. Because of this, we also recommend future research focus on developing more integrative correlates of plant mortality. For example, we may have captured more inter-specific variability in baseline hazards had we used a integrative measure of plant durability such as a trait that incorporates both wood density and stem architecture (REF).

Future research should also focus on examining how climate variables influence both growth-depedent and growth-independent hazards, particularly as climate-driven mortality is increasing globally (McDowell 2011, Get a Zepple paper). Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010. Unfortuantely, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3). However, based on anedoctal evidence, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by a 1998 El-Nino (REF). Furthermore, future research should also examine how different growth measures influence predictive accuracy of mortality models. We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive accuracy, and we are unsure how other common measures such as biomass growth (Chave's biomass paper) will compare.

@@

\matmethods{
\subsection*{Data}
We parametrized plant mortality models using individual growth and survival data \cite{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\textdegree;N, 79.85\textdegree;W). The climate on the island is warm throughout the year with a mean annual temperature of 26.1\textdegree and rainfall is seasonal with most of the 2500 mm falling between April and November. BCI is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI) and the permanent plot was established in 1980. Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in (REF).

Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter \cite{Condit:2012nz}. For the purpose of modelling mortality as a function of growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre. We also excluded individuals based on the following rules: 1) those that did not survive at least two censuses (two being required to estimate growth rate); 2) those that were not consistently measured at 1.3 m above ground; 3) those that were multi-stemmed; 4) those that resprouted or \textit{``returned from the dead''}; or 5) extreme outliers -- stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial diameter. We also discarded species that do not exhibit secondary growth (e.g. palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density.  In order to maintain statical independence of measurements, we then sub-sampled this dataset by randomly selecting one observation per individual out of a possible three censuses. In total, \Sexpr{n_inds(BCI_model_dataset_true_unique)} individuals of \Sexpr{n_spp(BCI_model_dataset_true_unique)} species were included in our analysis.

Wood density for each species was estimated by coring trees located within 15 km of the BCI plot, because increment borers are prohibited within the plot itself \cite{Wright:2010fl}. Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100\textdegree C) and dividing by the fresh volume (as measured by water displacement). Wood density was then calculated as an area-weighted average (g cm$^-3$), where area refers to the annulus represented by each piece, assuming a circular trunk.

@@

Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletion and bark drop) that can lead to trunk shrinkages overtime \cite{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative \cite{Clark:2007ve}. To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e. observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors \cite{Condit:2012nz}. This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period \cite{Condit:2012nz}. Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error. The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma(BCI_dbh_error_data)}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter. Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual (SUPP).

@@

For all models we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}. We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Gelman:1992ts, Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.

>>>>

\dropcap{R}ates of plant mortality are known to be highly variable among individuals, among coexisting species, between forests and from year-to-year [refs]. This variation has considerable consequences for forest structure and dynamics. Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools\cite{Chave:2005hz}, while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun. In models of forest dynamics, variation in mortality rates has been found to have a larger impact on forest structure than absolute growth rates\cite{Johnson-2016}. As such, improving our ability to understand the mortality process is therefore a priority for making skilful predictions about forest dynamics and carbon, water or nutrient cycles; especially in an era of rapid environmental change.

Two substantial challenges exist when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates (Poorter et al 2008). This requirement arises in part because the observable outcome -- alive vs. dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate). In classic survival analysis \cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$\cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
\end{equation}
Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required. As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys in long-term monitoring plots containining thousands of individuals and spanning several decades\cite{Anderson-Teixeira-2015a}.

The other major challenge when studying mortality -- similar to that faced in other fields like medicine and engineering -- is to derive suitable shapes for the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality. Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}. Generally, however, these influences have not been formalised as hazard functions\cite{Zens:2003jv}. Yet such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib;Zens:2003jv,McDowell:2011dr}. This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.

A particular challenge is to integrate the observed conflicting relationships between mortality and growth observed within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality than individuals from slower growing species\cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events\cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

@@

The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards\cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}. We show that regardless of growth measure, incorporating both hazards significantly improves model predictive accuracy. This is because the additive effects of both hazards explicitly allow models to account for two independent causes of plant mortality. The growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stressors (e.g. competition, parasitization, localized herbivory). By contrast, the growth-independent hazard accounts for deaths caused by stochastic events that arise irrespective of an individuals growth rate (e.g. windthrow).

Another advantage of partitioning mortality into growth-depedent and growth-independent hazards was that we could estimate the proportion of variation each attributed. Like many other studies\cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition an important factor influencing tropical tree demographic rates. Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted varability in mortality rates (Fig. \ref{fig5}), and thus, suggests that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.

Incorporating the effect of wood density on mortality rates also improved predictive performance. Our analyses revealed that the most parsimonious combination of wood density effects was when it was attributed to only growth-independent mortality. Specifically, we found that high wood density species had lower growth-independent mortality rates relative to their low wood density counterparts. This finding corroborating the observed negative correlation between mortality and wood density reported elsewhere\cite{Kraft:2010kq,Chave:2009iy}. More importantly, however, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.

While wood density effects are now being incorporated in mortality algorithms of many vegetation models\cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation. This means that such models are likely to severely underestimate the true varability in mortality rates. Consequently, because variation in mortality rates are known to have significant impacts on forest dynamics\cite{Johnson-2016}, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics, with flow on effects for ecosystem proporties such as GPP and NPP.

@@

The underestimation of inter-specific variation suggests that there are other species attributes besides wood density that affect tree mortality rates. It has been suggested that a species light demand and maximum size are two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}. While we did not explicitly model mortality as a function of these traits, our posthoc correlations  \ref{fig6} revealed that some inter-specific variation was indeed correlated with light demand. Specifically, we found that light demanding species (i.e. those with high gap index) were more susceptible to death due to both low growth and stochastic chance and required faster growth rates to achieve a given mortality rate relative to a species that readily recruits in shaded conditions, supporting findings by others\cite{Ruger:2011cv}. By contrast, we detected no correlations between a species maximum stem diameter and mortality rates, supporting findings by others\cite{Wright:2010fl}.

We recommend research should focus on examining how other traits influence growth-dependent and growth-independent hazards and what combination of traits minimises unexplained species variability. Researchers, however, should be mindful that the inclusion of additional traits comes at the cost of added complexity, increased uncertainty (i.e. more parameters to estimate) and increased minimum knowledge; all of which diminish the interpretability of models, but also the applicability of predicting mortality rates in many data poor ecosystems. Because of this, we also recommend future research focus on developing more integrative correlates of plant mortality. For example, we may have captured more inter-specific variability in baseline hazards had we used a integrative measure of plant durability such as a trait that incorporates both wood density and stem architecture.

Future research should also focus on examining how climate variables influence both growth-depedent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}. Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010. Unfortuantely, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3). However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by a El-Nino in 1997\textendash 1998. Droughts are known to have significant impacts on tree mortality rates in these ecosystems\cite{Condit:1995wl, Mcdowell:2013wf} Furthermore, future research should also examine how different growth measures influence predictive accuracy of mortality models. We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive accuracy, and we are unsure how other common measures such as biomass growth\cite{Chave:2005hz} will compare.

@@

\matmethods{
\subsection*{Data}
We parametrized plant mortality models using individual growth and survival data\cite{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\textdegree;N, 79.85\textdegree;W). The climate on the island is warm throughout the year with a mean annual temperature of 26.1\textdegree and rainfall is seasonal with most of the 2500 mm falling between April and November. BCI is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI) and the permanent plot was established in 1980.

Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter\cite{Condit:2012nz}. For the purpose of modelling mortality as a function of past growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre. Consequently, we modelled tree mortality as a function of past growth for censuses 1995\textendash 2000, 2000\textendash 2005 and 2005\textendash 2010. We also excluded individuals based on the following rules: 1) those that did not survive at least two censuses (two being required to estimate growth rate); 2) those that were not consistently measured at 1.3 m above ground; 3) those that were multi-stemmed; 4) those that resprouted or \textit{``returned from the dead''}; or 5) extreme outliers -- stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial diameter. We also discarded species that do not exhibit secondary growth (e.g. palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density.  In order to maintain statical independence of measurements, we then sub-sampled this dataset by randomly selecting one observation per individual out of a possible three censuses. In total, \Sexpr{n_inds(BCI_model_dataset_true_unique)} individuals of \Sexpr{n_spp(BCI_model_dataset_true_unique)} species were included in our analysis.

Wood density for each species was estimated by coring trees located within 15 km of the BCI plot, because increment borers are prohibited within the plot itself\cite{Wright:2010fl}. Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100\textdegree C) and dividing by the fresh volume (as measured by water displacement). Wood density was then calculated as an area-weighted average (g cm$^-3$), where area refers to the annulus represented by each piece, assuming a circular trunk.

In order to calculate our gap index (our measure of a species light dependence) we also used annual canopy census data collected during 1985\textendash 1990 and 1990\textendash 1995. The canopy census recorded the presence of vegetation in six height intervals, 0–2, 2–5, 5–10, 10–20, 20–30 and 30 m every 5-m across the 50 ha. We calculated the minimum observed number of vegetation strata above 2 m all 5x5 cells in each census period. We then transformed this to a binary gap index by where 1 refers to no vegetation above 2 m and 0 referring to densely vegetated by subtracting the above values from 5 (i.e., 0 = dense, 5 = gap). To acknowledge the influence of gaps on surrounding cells' light availability, we then calculated a weighted focal sum using a 3 x 3 cell moving window, for which the central cell was assigned a weight of 8, and the eight neighbouring cells were each assigned weights of 1. The combined impact of vegetation density of neighbouring cells on light availability at the central cell was therefore assumed to be equivalent to the impact of vegetation density at the central cell.

@@

Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletion and bark drop) that can lead to trunk shrinkages overtime\cite{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative\cite{Clark:2007ve}. To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e. observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors\cite{Condit:2012nz}. This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period\cite{Condit:2012nz}. Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error. The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma(BCI_dbh_error_data)}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter. Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual (SUPP).

@@

For all models we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}. We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic\cite{Gelman:1992ts, Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
